### 2015-06-11 ApresentacoesArtigos/quali/quali.tex
traduzida introdução

\usepackage{amssymb,amsmath,textcomp}
\usepackage[brazil,english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{braket}
\usepackage{booktabs}  %Possibilidade de linhas mais grossas nas tabelas.

@@

\begin{center} 
\LARGE{UNIVERSITY OF SÃO PAULO}  \\
\LARGE{SÃO CARLOS PHYSICS INSTITUTE}
\end{center}

@@

\centerline{\Huge{Complex networks for the participant}} 
\vspace{0.5cm}
%\centerline{\Huge{aplicada a reconhecimento de autoria}}

@@

\begin{singlespace}
Monograph presented to the Physics Graduate Program of the
São Carlos Institute of Physics, University of São Paulo,
for the qualification exam, as part of the requirements
for obtaining the title of Doctor in Sciences.\\

Concentration area: Applied Physics

Option: Computational Physics

Advisor: Prof. Dr. Osvaldo Novais de Oliveira Jr.
\end{singlespace}

@@

\centerline{\LARGE{{\bf ABSTRACT}}}

@@

\hspace*{-0.9cm} FABBRI, C. \textit{Complex networks for the participant}. 
São Carlos Institute of Physics. University of São Paulo, São Carlos, 2015.

@@

{\noindent
	Complex networks form one of the most active fields of recent physics.
	With respectable efforts for exhibiting advances to the general audience, 
	it seems, however, that few or none of these are targeted to the benefit of the individual that constitute such systems.
	That is, with a core knowledge about the field, and recipes for harnessing,
	provide means for the participant to interact and understand the networks he/she is in.
	This work aims to accomplish such task by means of the social networks of the participants.
	We verified that such networks exhibit time stability of topological measures and
	of basic connective sector sizes and exhibit differentiation of the textual production in each basic connective sector.
	We also formalized conceptualizations of these networks as OWL were they were possible, specially
	in relation to the social participation instances provided by law. Finally, software and
	data have been put available and used, as means to enable integrated analysis of different provenance and public benefit.
	Conceptual consequences have been documented and requires anthropological considerations.
	Furthermore, software, ontological and data contributions can be better documented and developed while
	a typological consideration of the physical properties observed in human interaction networks should
	bridge complex networks and the more traditional legacy of human sciences on the subject.
}

\vspace*{1.5cm}
\hspace*{-0.9cm} {\bf Keywords:} Complex networks. Social networks. Complexity. Anthropological physics. Linked data. Semantic web. Social participation. Text mining. Natural language processing.

@@

\chapter{Introduction}
Studies on human interaction networks have started long before modern computers, dating back to the nineteenth century, while the foundation of
social network analysis is generally attributed to the psychiatrist Jacob Moreno in mid twentieth century~\cite{newmanBook}.
With the increasing availability of data related to human interactions, research on these networks has grown continuously.
Contributions can now be found in a variety of fields, from social sciences and humanities~\cite{latour2013} to computer science~\cite{bird} and physics~\cite{barabasiHumanDyn,newmanFriendship}, given the multidisciplinary nature of the topic.
One of the approaches from an exact science perspective is to represent interaction networks as complex networks~\cite{barabasiHumanDyn,newmanFriendship}, with which 
several features of human interaction have been revealed.
For example, the topology of human interaction networks exhibits a scale-free trace, which points to the existence of a small number of highly connected hubs and a large number of poorly connected nodes.

There is gap of knowledge and technology between the complex networks legacy and the usufruct of the participant. This hiatus is reactive, and events suggest that it will maintain itself as an ecosystem of knowledge, technology and undertake of society in all scales. It should ease for example: entrepreneur goals, elaboration and preparation of documents, rapid achievement of knowledge. In general: processes that refer to collection and diffusion of goods and information.
% Argumentar que é tido que o correto entendimento destas estruturas
% será mais forte que XXX (ver no curso da Melanie/CE).

This work presets a confirmation of such scenario and advances.
Some strategies were selected to verify the complex networks to the
advantage of the participant. Specially, very simple experiments
seems able to modify social structures. In this context, we verified
temporal stabilities in human interaction networks, and exposed that
primitive sectors of the networks (hubs, intermediary and periphery)
produce texts that are very different. This is useful for a
not stigmatizing typology of participants in interaction networks
via quantitative criteria. This yielded audiovisualization and
interconnection of data as art and technological gadgets, 
as support for scientific research. Applications were complemented
with the (Brazilian) Presidency of the Republic and UNDP.

Next section presents directions on related work. Section~\ref{sec:mat} is dedicated to the data analyzed. Section~\ref{sec:met} holds the methods used to reach results. Results are given in Section~\ref{sec:res}. The chronogram and outline of finished, ongoing and planed tasks are in Section~\ref{sec:chr}. Conclusions are in Section~\ref{sec:con} followed by acknowledgments and the bibliography.
% Interacting with own network, with constant feedback, study of the self
% Annotation of own data, study of them. Extend analysis within the route.
%In this work, we present advances in handing the participant ways to explore and harness the social networks he/she participates in.
%This is not an easy task, therefore we selected

\section{Literature review}
The field  of complex networks is relatively new ($\approx 25$ years)
and literature presents diverging definitions.
One definition that is having increasing acceptance
considers a complex network a ``large graph with non-trivial topological features''.
This definition is misleading in at least two points: there are networks of interest with trivial topological features, such as the paradigmatic Erdös-Rényi network and the lattice network~\cite{newman}. 
Second, it fails to deliver the fundamental message that the complex network is not an isolated mathematical graph structure. Complex networks of interest are real networks or idealized models for understanding them. Besides that, not only large graphs are of interest, but small graphs are very often used as toy examples and measured as extension of larger structures.
A definition, still far from perfect, but preferred in this work, is:
%``often large graphs considered in the physical
%or natural environment they reside''.
``often large and non-trivial graphs considered in,
or for the consideration of, 
the environment they reside''.
This definition resolves both issues.

@@

\chapter{Materials}
\section{The Gmane public database of email lists (benchmarks)}
\section{Facebook, Twitter, Participa.br, Cidade Democrática, AA}
\section{My own social networks}
Considerations about the right to annotate.

>>>>

\usepackage{amssymb,amsmath,textcomp}
\usepackage[brazil]{babel}
\usepackage[utf8]{inputenc}
\usepackage{braket}
\usepackage{booktabs}  %Possibilidade de linhas mais grossas nas tabelas.

@@

\begin{center} 
%\LARGE{UNIVERSITY OF SÃO PAULO}  \\
%\LARGE{SÃO CARLOS PHYSICS INSTITUTE}
\LARGE{UNIVERSIDADE DE SÃO PAULO}  \\
\LARGE{INSTITUTO DE FÍSICA DE SÃO CARLOS}
\end{center}

@@

%\centerline{\Huge{Complex networks for the participant}} 
\centerline{\Huge{Redes complexas para o participante}} 
\vspace{0.5cm}
%\centerline{\Huge{aplicada a reconhecimento de autoria}}

@@

\begin{singlespace}
Monografia apresentada ao Programa de Pós-Graduação em Física 
do Instituto de Física de São Carlos da Universidade de São Paulo, para o Exame de Qualifica\c c\~ao 
como parte dos requisitos para obten\c c\~ao do t\'itulo de Doutor em Ci\^encias.\\

Área de concentração: Física Básica

Orientador: Prof. Dr. Osvaldo Novais de Oliveira Jr.
%Monograph presented to the Physics Graduate Program of the
%São Carlos Institute of Physics, University of São Paulo,
%for the qualification exam, as part of the requirements
%for obtaining the title of Doctor in Sciences.\\
%
%Concentration area: Applied Physics
%
%Option: Computational Physics
%
%Advisor: Prof. Dr. Osvaldo Novais de Oliveira Jr.
\end{singlespace}

@@

%\centerline{\LARGE{{\bf ABSTRACT}}}
\centerline{\LARGE{{\bf RESUMO}}}

@@

%\hspace*{-0.9cm} FABBRI, C. \textit{Complex networks for the participant}. 
\hspace*{-0.9cm} FABBRI, C. \textit{Redes complexas para o participante}. 
Instituto de Física de São Carlos, Universidade de São Paulo, São Carlos, 2015. 
%São Carlos Institute of Physics. University of São Paulo, São Carlos, 2015.

@@

{\noindent
	As redes complexas formam uma das áreas mais ativas da física redente.
	Há esforços consideráveis para apresentar estes avanços ao público geral, mas tudo indica
	que poucos ou nenhum são voltados para o benefício do indivíduo que constitui estes sistemas.
	Ou seja, com um núcleo de conhecimento da área, e receitas para aproveitamento,
	fornece meios para o participante interagir e entender as redes nas quais ele se encontra.
	Este trabalho objetiva realizar tal tarefa por meio das redes sociais do participante.
	Verificamos que tais redes exibem uma estabilidade temporal de medidas temporais e dos
	tamanhos relativos dos setores conectivos básicos (hubs, intermediários, periféricos).
	Observamos uma diferenciação da produção de texto de cada setor básico.
	Também formalizamos as conceitualizações destas redes como OWL onde foi possível,
	principalmente as instâncias de participação dosial previstas por lei e implementadas como software.
	Software e dados foram disponibilizados e usados. Protocolos escolhidos para
	facilitar a integração de estruturas de diferentes procedências,
	para reutilização dos dados em outros trabalhos e pesquisas, e para o benefício público.
	Consequências conceituais requerem considerações antropológocas e estão sendo documentadas.
	Próximos passos são: melhor documentação e desenvolvimento do aparato em software, ontologias e dados;
	considerações tipológicas das propriedades físicas observadas nas redes de interação humana, com
	atenção aos outliers, às relações entre topologia do agente e texto produzido, e à ponte com
	a bagagem mais tradicional das ciências humanas no assunto.
}

%{\noindent
%	Complex networks form one of the most active fields of recent physics.
%	With respectable efforts for exhibiting advances to the general audience, 
%	it seems, however, that few or none of these are targeted to the benefit of the individual that constitute such systems.
%	That is, with a core knowledge about the field, and recipes for harnessing,
%	provide means for the participant to interact and understand the networks he/she is in.
%	This work aims to accomplish such task by means of the social networks of the participants.
%	We verified that such networks exhibit time stability of topological measures and
%	of basic connective sector sizes and exhibit differentiation of the textual production in each basic connective sector.
%	We also formalized conceptualizations of these networks as OWL were they were possible, specially
%	in relation to the social participation instances provided by law. Finally, software and
%	data have been put available and used, as means to enable integrated analysis of different provenance and public benefit.
%	Conceptual consequences have been documented and requires anthropological considerations.
%	Furthermore, software, ontological and data contributions can be better documented and developed while
%	a typological consideration of the physical properties observed in human interaction networks should
%	bridge complex networks and the more traditional legacy of human sciences on the subject.
%}

\vspace*{1.5cm}
%\hspace*{-0.9cm} {\bf Keywords:} Complex networks. Social networks. Complexity. Anthropological physics. Linked data. Semantic web. Social participation. Text mining. Natural language processing.
\hspace*{-0.9cm} {\bf Palavras-chave:} Redes complexas. Redes sociais. Complexidade. Física antropológica.
Dados ligados. Web semântica. Participação social. Mineração de texto. Processamento de linguagem natural.

@@

%\chapter{Introduction}
\chapter{Introdução}
Estudos sobre redes de interação humana foram iniciados bem antes dos computadores modernos,
datam do século XIX, enquanto a fundação da ``Análise de Redes Sociais''/ARS (\emph{Social Network Analysis}/SNA)
é geralmente atribuida ao psiquiatra Jacob Moreno na metade do século vinte~\cite{newmanBook}.
Com a crescente disponibilidade de dados relacionados à interação humana, a pesquisa destas redes tem aumentado continuamente.
Contribuições podem ser encontradas em uma variedade de áreas, de ciências sociais e humanidades~\cite{latour2013}
a ciências sociais~\cite{bird} e física~\cite{barabasiHumanDyn,newmanFriendship},
dada a natureza multidisciplinar do assunto.
Uma das abordagens da perspectiva de uma ciência exata é representar a rede de interação como uma rede
complexa~\cite{barabasiHumanDyn,newmanFriendship},
com a qual algumas características foram reveladas.
Por exemplo, a topologia das redes de interação humana exibem um traço livre de escala,
o que aponta para a existência de um pequeno número de hubs super conectados
e um grande número de vértices pouco conectados.

Há um hiato de conhecimento e tecnologi entre o legado de redes complexas e o usufruto do participante.
Este hiato é reativo, e há evidência de que conseguirá se manter como um ecossistema de conhecimento, tecnologia
e empreendimento da sociedade em todas as suas escalas.
Deve facilitar, por exemplo:
elaboração e preparação de documentos, aquisição rápida de conhecimento,
realização de empreitadas coletivas.
Em geral: processos de coleta e difusão de informação (e bens).
% Argumentar que é tido que o correto entendimento destas estruturas
% será mais forte que XXX (ver no curso da Melanie/CE).

Este trabalho apresenta uma confirmação deste cenário e avanços.
Algumas estratégias foram selecionadas para verificar a aplicabilidade
de conceitos de redes complexas para o benefício do participane.
Em especial, experimentos muito simples parecem capazes de modificar
estruturas sociais. Neste contexto, verificamos estabilidades temporais
nas redes de interação humana, e expomos que os setores primitivos
das redes (hubs, intermediários e periféricos)
produzem textos bastante diferentes entre si.
Este conhecimento é útil para uma tipologia não estigmatizante
de participantes em redes de interação.
A audiovisualização e interconexão de dados com arte e engenhocas
em software deram suporte contínuo à pesquisa científica.
Aplições foram complementadas com a Presidência da República
e o PNUD/ONU.

A próxima seção apresenta considerações gerais sobre a literatura.
A Seção~\ref{sec:mat} é dedicada aos dados analizados.
A Seção~\ref{sec:met} contém os métodos usados para atingir os resultados, que são explicitados na Seção~\ref{sec:res}. O cronograma de atividades e uma comparação entre afazeres planejados, em andamento e finalizados
estão na Seção~\ref{sec:chr}.
A monografia termina com as conclusões na Seção~\ref{sec:con}, seguida de agradecimentos e bibliografia.

% Interacting with own network, with constant feedback, study of the self
% Annotation of own data, study of them. Extend analysis within the route.
%In this work, we present advances in handing the participant ways to explore and harness the social networks he/she participates in.
%This is not an easy task, therefore we selected

\section{Literature review}
The field  of complex networks is relatively new ($\approx 25$ years)
and literature presents diverging definitions for the field itself.
One definition that is having increasing acceptance
considers a complex network a ``large graph with non-trivial topological features''.
This definition is misleading in at least two points: there are networks of interest with trivial topological features, such as the paradigmatic Erdös-Rényi network and the lattice network~\cite{newman}. 
Second, it fails to deliver the fundamental message that the complex network is not an isolated mathematical graph structure. Complex networks of interest are real networks or idealized models for understanding them. Besides that, not only large graphs are of interest, but small graphs are very often used as toy examples and measured as extension of larger structures.
A definition, still far from perfect, but preferred in this work, is:
%``often large graphs considered in the physical
%or natural environment they reside''.
``often large and non-trivial graphs considered in,
or for the consideration of, 
the environment they reside''.
This definition resolves both issues.

@@

\chapter{Materials}\label{sec:mat}
\section{The Gmane public database of email lists (benchmarks)}
\section{Facebook, Twitter, Participa.br, Cidade Democrática, AA}
\section{My own social networks?}
Considerations about the right to annotate.
